Concurrent Programming with CXL

### 2020-06-14 book/paper.tex
Updates
Fixes

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface using suspension.}
\label{fig:suspension}
\end{figure}

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}.
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variabled.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
There are two operations on condition variables: \texttt{wait} and \texttt{signal}.
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

\texttt{wait} releases the monitor lock and blocks while trying
to procure the condition variable (which is a semaphore that is 0 when
\texttt{wait} is invoked).
\texttt{signal} vacates the condition variable, passing the baton to
a process trying to procure it, and then procures the monitor lock.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
\texttt{V}.

@@

As in Hoare monitors, there is a lock associated with each monitor,
and the lock is acquired upon entry to a method and released upon exit.
Mesa monitors also have condition variable that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation release the monitor
lock.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
he baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point $q$ gets the baton,
assuming there even was such a process $q$.

@@

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and let's that process into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When process $p$ leaves the bathroom,
if also checks each of the bedrooms to see if there is a process $q$ that
can run.  But instead of letting $q$ go straight into the bathroom, $q$
goes into the hall, joining any other processes that are also waiting to
enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
processes in the hall is let into the bathroom.  But that may not be $q$.
Assuming every process eventually leaves the bathroom and the system is
fair, eventually $q$ will enter the bathroom.  However, it is no longer
guaranteed that its condition holds.  Therefore, the first thing $q$ must
do is check to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.

It turns out that with the Mesa monitor there is another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Note that called \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  Mesa also has a function called
\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

@@

>>>>

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}.
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
There are two operations on condition variables: \texttt{wait} and \texttt{signal}.
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

\texttt{wait} releases the mutex and blocks while trying
to procure the condition variable (which is a semaphore that is 0 when
\texttt{wait} is invoked).
\texttt{signal} vacates the condition variable, passing the baton to
a process trying to procure it, and then procures the mutex.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
\texttt{V}.

@@

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variable that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation release the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
the baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point $q$ gets the baton,
assuming there even was such a process $q$.

@@

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let in a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and lets that process into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When process $p$ leaves the bathroom,
if also checks each of the bedrooms to see if there is a process $q$ that
can run.  But instead of letting $q$ go straight into the bathroom, $q$
goes into the hall, joining any other processes that are also waiting to
enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
processes in the hall is let into the bathroom.  But that may not be $q$.
Assuming every process eventually leaves the bathroom and the system is
fair, eventually $q$ will enter the bathroom.  However, it is no longer
guaranteed that its condition holds.  Therefore, the first thing $q$ must
do is check, again, to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also has a function called
\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

@@

The \texttt{synch} module supports Mesa condition variables.  However, like C, Java, and Python,
it does not have built-in language support for them but instead associates condition variables
with explicit locks.
Figure~\ref{fig:cv} shows the \texttt{synch} interface to condition variables.
\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
to a lock \texttt{lk}.
The condition variable itself is represented by a dictionary containing the pointer to the lock
and a bag of name tags of processes waiting on the condition variable.
(The \texttt{synchS} library instead has a list of contexts.)

\texttt{wait} adds the nametag of the process to the bag.  This increments the number of processes
in the bag with the same context.  \texttt{wait} then waits until that count is restored to the
value that it had upon entry to \texttt{wait}.  \texttt{notify} removes an arbitrary context from
the bag, allowing one of the processes with that context to return from \texttt{wait}.
\texttt{notifyAll} empties out the entire bag, allowing all processes in the bag to resume.
